---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# LidarLDA

<!-- badges: start -->
<!-- badges: end -->

The goal of LidarLDA is to fit the a modified version of the Latent Dirichlet Allocation (LDA) model to LIDAR data. Ultimately, LDA will identify clusters with cluster-specific absorptance probabilities and will characterize each pixel in relation to the relative abundance of these clusters.

## Installation

You can install this package from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("drvalle1/LidarLDA")
```
## Example

This is a basic example which shows you to run LidarLDA:

```{r example}
library(LidarLDA)

## basic example code
P=1000
H=10

Model.Results=LidarLDA(y=matrix(1,P,H),
                       n=matrix(2,P,H),
                       ncomm=10,
                       a.phi=1,b.phi=1,
                       gamma=0.1,ngibbs=100,
                       nburn=50,theta.post=F,phi.post=F)
```
